\section{Evaluation}
\label{sec:evaluation}

In this section we evaluate the design from \S\ref{sec:design}, specifically our implementation from \S\ref{sec:implementation}. We show how the implementation performs on the example from \S\ref{sec:introduction} in \S\ref{sec:eval:introduction}, on microbenchmarks in \S\ref{sec:eval:overhead}, and then on real projects that currently use \Make{} -- namely FSATrace (\S\ref{sec:eval:fsatrace}), Redis (\S\ref{sec:eval:redis}), Vim (\S\ref{sec:eval:vim}), Tmux (\S\ref{sec:eval:tmux}), and Node.js (\S\ref{sec:eval:node}). For larger projects we look at both whether the dependencies are correct and the performance.

For benchmarks, the first three (\S\ref{sec:eval:introduction}, \S\ref{sec:eval:fsatrace} and \S\ref{sec:eval:redis}) were run on a 4 core Intel i7-4790 3.6GHz CPU, with 16Gb of RAM. The remaining benchmarks were run on a 32 core Intel Xeon E7-4830 2.13GHz, with 62Gb of RAM.

\subsection{Validating the claims from \S\ref{sec:introduction}}
\label{sec:eval:introduction}

In \S\ref{sec:introduction} we claimed that the following build script is ``just as good'' as a proper \Make script.

\begin{verbatim}
gcc -c main.c
gcc -c util.c
gcc -o main.exe main.o util.o
\end{verbatim}

There are two axes on which to measure ``just as good'' -- correctness and performance. Performance can be further broken down into how much rebuilds, how much parallelism can be achieved and how much overhead there is.

\paragraph{Correctness} \Rattle is correct, in that the reference semantics is running all the commands, and as we have shown in \S\ref{sec:design} and \S\ref{sec:proof}, and tested for with examples, \Rattle obeys those semantics. In contrast, the \Make version may have missing dependencies which causes it not to rebuild. Examples of failure to rebuild include if \texttt{gcc} changes or any headers used but not listed in the \Make script change.

\paragraph{Rebuilding too much} \Rattle only rebuilds a command if some of the files it reads or writes have changed. It is possible that a command only depends on part of a file, but at the level of abstraction \Rattle works, it never rebuilds too much. As a matter of implementation, \Rattle detects changes by hashing the file contents, while \Make uses the modification time. As a consequence, if a file is modified, but it's contents do not change (e.g. using \texttt{touch}), \Make will rebuild but \Rattle will not.

\paragraph{Parallelism} The script from \S\ref{sec:introduction} has three commands -- the first two can run in parallel, while the the third must wait for both to finish. \Make is given all this information by dependencies, and will always achieve as much parallelism as possible. In contrast, \Rattle has no such knowledge, so has to recover the parallelism by speculation (see \S\ref{sec:speculation}). During the first execution, \Rattle has no knowledge about even which commands are coming next (as described in \S\ref{sec:monadic}), so has no choice but to execute each command serially, with less parallelism than \Make. In subsequent executions \Rattle uses speculation to always speculate on the second command (as it never has a hazard with the first), but never speculate on the third until the first two have finished (as they are known to conflict). Interestingly, sometimes \Rattle executes the third command (because it got to that point in the script), and sometimes it speculates it (because the previous two have finished) -- it is a race condition where both alternatives are equivalent. While \Rattle has less parallelism on the first execution, by using shared storage for speculation traces, that can be reduced to the first execution \emph{ever}, rather than the first execution for a given user.

\paragraph{Overhead} The overhead inherent in \Rattle is greater than that of \Make, as it hashes files, traces command execution, computes potential hazards, figures out what to speculate and writes to a shared cloud store. To measure the overhead, and prove the other claims in this section, we created a very simple pair of \texttt{main.c} and \texttt{util.c} files where \texttt{main.c} calls \texttt{printf} using a constant returned by a function in \texttt{util.c}. We then measured the time to do:

\begin{enumerate}
\item An initial build from a clean checkout.
\item A rebuild when nothing had changed.
\item A rebuild with whitespace changes to \texttt{main.c}, which result in the same \texttt{main.o} file being produced.
\item A rebuild with meaningful changes to \texttt{main.c}, resulting in a different \texttt{main.o} file.
\item A rebuild with meaningful changes to both C files.
\end{enumerate}

We performed the above steps with 1, 2 and 3 threads, on Linux. To make the parallelism obvious, we modified \texttt{gcc} to sleep for 1 second before starting. The numbers are:

% RAW RESULTS
% $ rattle-benchmark intro
%   make   -j1: 3.35s 0.00s 2.19s 2.19s 3.28s
%   rattle -j1: 3.28s 0.00s 1.12s 2.16s 3.29s
%   make   -j2: 2.18s 0.00s 2.19s 2.23s 2.22s
%   rattle -j2: 3.28s 0.00s 1.12s 2.21s 2.20s
%   make   -j3: 2.20s 0.00s 2.19s 2.19s 2.19s
%   rattle -j3: 3.36s 0.00s 1.10s 2.22s 2.20s
%   make   -j4: 2.20s 0.00s 2.19s 2.18s 2.19s
%   rattle -j4: 3.28s 0.00s 1.14s 2.21s 2.20s

\vspace{3mm}
\begin{tabular}{l|r|r||r|r||r|r}
Number of threads & \multicolumn{2}{c||}1 & \multicolumn{2}{c||}2 & \multicolumn{2}{c}3 \\
Tool & \Make & \Rattle & \Make & \Rattle & \Make & \Rattle \\
\hline
1) Initial build & 3.35s & 3.28s & 2.18s & 3.28s & 2.20s & 3.36s \\
2) Nothing changed & 0.00s & 0.00s & 0.00s & 0.00s & 0.00s & 0.00s \\
3) \texttt{main.c} changed whitespace & 2.19s & 1.12s & 2.19s & 1.12s & 2.19s & 1.10s \\
4) \texttt{main.c} changed & 2.19s & 2.16s & 2.23s & 2.21s & 2.19s & 2.22s \\
5) Both C files changed & 3.28s & 3.29s & 2.22s & 2.20s & 2.19s & 2.20s \\
\end{tabular}
\vspace{3mm}

As expected, we see that in the initial build \Rattle doesn't exhibit parallelism, but \Make can (1). In contrast, \Rattle can benefit when a file changes in whitespace only and the resulting object file doesn't change, while \Make can't (3). We see 3 threads has no benefit over 2 threads, as this build contains no more parallelism opportunities. Comparing the non-sleep portion of the build, \Make and \Rattle are quite evenly matched, typically within a few milliseconds, showing low overheads. We focus on the overheads in the next section.

\subsection{Measuring overhead}
\label{sec:eval:overhead}

In order to determine what overhead \Rattle introduces, we ran a fixed set of commands with increasingly more parts of \Rattle enabled. \Rattle command execution builds on the command execution from \Shake \cite{shake}, which in turn uses \Fsatrace for tracing and the Haskell \texttt{process} library for command execution. Therefore, we ran the commands in a clean build directory in 7 different ways:

\begin{enumerate}
\item Using \texttt{make -j1}, as a baseline.
\item Using \texttt{System.Process} from the Haskell \texttt{process} library.
\item Using \texttt{cmd} from the Haskell \Shake library \cite{shake}, which builds on top of the \texttt{process} library.
\item Using \texttt{cmd} from \Shake, but wrapping the command with \Fsatrace for file tracing.
\item Using \texttt{cmd} from \Shake with the \texttt{Traced} setting, which runs \Fsatrace and collects the results.
\item Using \Rattle with no speculation or parallelism, and not storing any results to shared storage.
\item Using \Rattle with all features turned on, including shared storage.
\end{enumerate}

To obtain a set of commands typical of building, we took the latest version of \Fsatrace\footnote{\url{https://github.com/jacereda/fsatrace/commit/41fbba17da580f81ababb32ec7e6e5fd49f11473}} and ran \texttt{make -j1}, capturing the commands that were executed. On Windows \Fsatrace runs 25 commands (21 compiles, 4 links). On Linux \Fsatrace runs 9 commands (7 compiles, 2 links). On Linux the list of commands produces write-write hazards, because it compiles some files (e.g. \texttt{shm.c}) twice, once with \texttt{-fPIC} (position independent code), and once without. However, both times it passes \texttt{-MMD} to cause \texttt{gcc} to produce \texttt{shm.d} which is used for dependencies -- we removed the \texttt{-MMD} flag as it doesn't impact the benchmark. We ran all sets of commands five times, and took the average of the three fastest, on both Windows and Linux.

% RAW RESULTS (final number is avg of fastest 3, to ignore swapping etc)
% $ rattle-benchmark micro
%   WINDOWS (with Windows Defender)
%     make: 13.14s 10.47s 11.04s 11.67s 11.86s = 11.06s
%     System.Process: 12.28s 12.41s 12.90s 13.07s 13.24s = 12.53s
%     shake.cmd: 13.59s 13.83s 13.88s 13.99s 14.08s = 13.77s
%     shake.cmd fsatrace: 16.54s 17.06s 16.55s 16.41s 16.39s = 16.45s
%     shake.cmd traced: 16.52s 16.42s 16.84s 16.69s 16.70s = 16.54s
%     rattle: 20.70s 18.51s 18.93s 18.50s 18.51s = 18.51s
%     rattle share: 18.94s 19.10s 18.74s 18.74s 18.63s = 18.70s
%   WINDOWS (no Windows Defeneder)
%     make: 12.31s 9.73s 9.98s 14.97s 10.16s = 9.96s
%     System.Process: 10.20s 10.25s 10.32s 10.40s 10.44s = 10.26s
%     shake.cmd: 10.53s 10.68s 10.62s 10.59s 10.70s = 10.58s
%     shake.cmd fsatrace: 12.68s 12.58s 12.85s 12.72s 12.80s = 12.66s
%     shake.cmd traced: 13.31s 12.99s 12.86s 13.90s 16.26s = 13.06s
%     rattle: 16.59s 14.51s 14.48s 14.82s 14.30s = 14.43s
%     rattle share: 14.56s 14.48s 14.62s 14.62s 14.56s = 14.53s
%   LINUX
%     make: 1.26s 1.22s 1.19s 1.23s 1.17s = 1.19s
%     System.Process: 1.19s 1.19s 1.22s 1.24s 1.16s = 1.18s
%     shake.cmd: 1.20s 1.18s 1.17s 1.21s 1.18s = 1.17s
%     shake.cmd fsatrace: 1.26s 1.20s 1.26s 1.25s 1.25s = 1.23s
%     shake.cmd traced: 1.25s 1.23s 1.27s 1.21s 1.23s = 1.23s
%     rattle: 1.50s 1.29s 1.25s 1.25s 1.25s = 1.25s
%     rattle share: 1.31s 1.31s 1.31s 1.26s 1.24s = 1.27s

\vspace{3mm}
\begin{tabular}{l|rrr|rrr}
Commands & \multicolumn{3}{c|}{Windows} & \multicolumn{3}{c}{Linux} \\
\hline
1) Make                      &  9.96s & 100\% &       &    1.19s & 100\% & \\
2) process                   & 10.26s & 103\% &  +3\% &    1.18s &  99\% & -1\% \\
3) \Shake                    & 10.58s & 106\% &  +3\% &    1.17s &  98\% & -1\% \\
4) \Shake + \Fsatrace        & 12.66s & 127\% & +21\% &    1.23s & 103\% & +5\% \\
5) \Shake + \texttt{Traced}  & 13.06s & 131\% &  +4\% &    1.23s & 103\% & +0\% \\
6) \Rattle                   & 14.43s & 145\% & +14\% &    1.25s & 105\% & +2\% \\
7) \Rattle + everything      & 14.53s & 146\% &  +1\% &    1.27s & 107\% & +2\% \\
\end{tabular}
\vspace{3mm}

Both Windows and Linux have three columns -- the time taken (average of the three fastest of five runs), that time as a percentage of the \Make run, and the delta from the row above. The results are significantly different between platforms:

\paragraph{Windows} On Windows, we see that the total overhead of \Rattle makes the execution 46\% slower. Of the parts, 21\% of the slowdown is from \Fsatrace (due to hooking Windows kernel API), with the next greatest overhead being from \Rattle itself. Of the \Rattle overhead, the greatest slowdown is caused by canonicalizing file names. Using the default NTFS file system, Windows considers paths to be case insensitive. As a result, we observe paths like \verb"C:\windows\system32\KERNELBASE.dll", which on disk are called \verb"C:\Windows\System32\KernelBase.dll", but can also be accessed by names such as \verb"C:\Windows\System32\KERNEL~1.DLL". Unfortunately, Windows also supports case sensitive file systems, so simply using case-insensitive equality is insufficient.

On Windows, enabling the default anti-virus (Windows Defender) has a significant impact on the result, increasing the \Make baseline by 11\% and the final time by 29\%. These results were collected with the anti-virus disabled.

\paragraph{Linux} In contrast, on Linux, the total overhead is only 7\%, of which nearly all (5\%) comes from tracing.

\postparagraphs

These results show that tracing has a minor but not insignificant effect on Linux, whereas on Windows can be a substantial performance reduction. Additionally, the Windows results had far more variance, in particular large outliers. As a consequence, our remaining benchmarks were all performed on Linux.

\subsection{\Fsatrace}
\label{sec:eval:fsatrace}

To compare \Make and \Rattle on \Fsatrace we took the commands we extracted for \S\ref{sec:eval:overhead} and ran the build script for the 100 previous commits in turn, starting with a clean build then performing incremental builds. To make the results readable, we hid any commands where all versions were < 0.02s, resulting in 26 interesting commits. We ran with 1 to 4 threads, but omit the 2 and 3 thread case as they typically fall either on or just above the 4 thread results.

\vspace{3mm}

\begin{tikzpicture}
\begin{axis}[
  title={Compile time at each successive commit},
  width=\textwidth,
  height=5cm,
  ylabel={Seconds},
  ymin=0,
  xmin=0,
  xmax=15,
]
\addplot [color=cyan, mark=o] table [x expr=\coordindex, y=make1] {data/fsatrace.dat};    \addlegendentry{\Make -j1}
\addplot [color=cyan, mark=*] table [x expr=\coordindex, y=make4] {data/fsatrace.dat};    \addlegendentry{\Make -j4}
\addplot [color=purple, mark=triangle] table [x expr=\coordindex, y=rattle1] {data/fsatrace.dat};  \addlegendentry{\Rattle -j1}
\addplot [color=purple, mark=triangle*] table [x expr=\coordindex, y=rattle4] {data/fsatrace.dat};  \addlegendentry{\Rattle -j4}
\end{axis}
\end{tikzpicture}

As we can see, the first build is always > 1s for \Rattle, but \Make is able to optimise it as low as 0.33s with 4 threads. Otherwise, \Rattle and \Make are competitive -- users would struggle to see the difference. The one commit that does show some variation is commit 2, where both are similar for 1 thread, but \Rattle at 4 threads is slower than 1 thread, while \Make is faster. The cause is speculation leading to a write-write hazard. Concretely, the command for linking \texttt{fsatrace.so} changed to include a new file \texttt{proc.o}. \Rattle starts speculating on the old link command, then gets the command for the new link. Both commands write to \texttt{fsatrace.so}, leading to a hazard, and causing \Rattle to restart without speculation.

% WriteWriteHazard /tmp/extra-dir-39079319872788/fsatrace.so Cmd [EchoStderr False] ["cc","-shared","src/unix/fsatraceso.os","src/emit.os","src/unix/shm.os","-o","fsatrace.so","-ldl","-lrt"] Cmd [EchoStderr False] ["cc","-shared","src/unix/fsatraceso.os","src/emit.os","src/unix/shm.os","src/unix/proc.os","-o","fsatrace.so","-ldl","-lrt"] Restartable

\subsection{Redis}
\label{sec:eval:redis}

\vspace{2mm}
\begin{tikzpicture}
\begin{axis}[
  title={Compile time at each successive commit},
  width=\textwidth,
  height=5cm,
  ylabel={Seconds},
  ymin=0,
  xmin=0,
  xmax=30,
]
\addplot [color=cyan, mark=o] table [x expr=\coordindex, y=make1] {data/redis.dat}; 	\addlegendentry{\Make -j1}
\addplot [color=cyan, mark=*] table [x expr=\coordindex, y=make4] {data/redis.dat}; 	\addlegendentry{\Make -j4}
\addplot [color=purple, mark=triangle] table [x expr=\coordindex, y=rattle1] {data/redis.dat}; 	\addlegendentry{\Rattle -j1}
\addplot [color=purple, mark=triangle*] table [x expr=\coordindex, y=rattle4] {data/redis.dat}; 	\addlegendentry{\Rattle -j4}
\end{axis}
\end{tikzpicture}

Redis is an in-memory NoSQL database written in C, developed over the last 10 years. Redis is built using recursive \Make \cite{miller:recursive_make}, where the external dependencies of Redis (e.g. Lua and Linenoise) have their own \Make scripts which are called by the main Redis \Make script. Using stamp files, on successive builds, the external dependencies are not checked -- meaning that users must clean their build when external dependencies change.

When benchmarking \Rattle vs \Make we found that \Rattle was approximately 10\% slower due to copying files to the cloud cache, something \Make does not do. Therefore, for a fairer comparison, we disabled the cloud cache. For local builds, the consequence of disabling the cloud cache is that if a file changes, then changes back to a previous state, we will have to rebuild rather than get a cache hit -- something that never happens in this benchmark. With those changes \Rattle is able to perform very similarly to \Make, even though it properly checks the dependencies. Differences occur on the initial build, where \Rattle has no information to speculate with, and commit 1, where \Rattle correctly detects no work is required but \Make rebuilds some dependency information.

\begin{comment}
% I reimplemented Stack in Rattle. Not sure it's useful given how much other evaluation stuff we have.
\subsection{Reimplementing Stack}

\Rattle assumes that each command is atomic - it cannot be subdivided into smaller parts. If a command is secretly two independent commands then they should usually be expressed as such so they can be individually skipped.

Compound commands: Sometimes a command will produce something that is user specific (not great for caching), but the next step will remove the user specificity (good for caching). To fix that we allow compound commands, by conjoining two commands with \texttt{\&\&}. Sometimes the sole purpose of the second command can be to strip machine-unique data from the first command.

As another example, the GHC package database has additional entries added every time a package is installed, making the output a consequence of the original file\footnote{As a consequence many build systems, including \Bazel and \Rattle, use multiple package databases with only one entry per database}.

and some memoisation operations (\texttt{memo})
\end{comment}

\subsection{Vim}
\label{sec:eval:vim}

\begin{tikzpicture}
\begin{axis}[
  title={Compile time at each successive commit},
  width=\textwidth,
  height=5cm,
  ylabel={Seconds},
  ymin=0,
  xmin=0,
  xmax=39,
]

\addplot [color=cyan, mark=o] table [x expr=\coordindex, y=make1] {data/vim.dat}; 	\addlegendentry{\Make -j1}
\addplot [color=cyan, mark=*] table [x expr=\coordindex, y=make4] {data/vim.dat}; 	\addlegendentry{\Make -j4}
\addplot [color=purple, mark=triangle] table [x expr=\coordindex, y=rattle1_noshared] {data/vim.dat}; 	\addlegendentry{\Rattle -j1}
\addplot [color=purple, mark=triangle*] table [x expr=\coordindex, y=rattle4_noshared] {data/vim.dat}; 	\addlegendentry{\Rattle -j4}
\end{axis}
\end{tikzpicture}

Vim is a popular text editor, over 25 years old. The majority of the source code is Vim script and C, and it is built with \Make. To compare the original \Make based build to \Rattle, we built Vim over a series of commits, using both the original \Make and a \Rattle script created by recording the commands \Make executed on a clean build. The benchmarks show that \Make and \Rattle had similar performance for almost every commit, apart from two.

\begin{enumerate}
\item For the 20th commit, \Make did much more work than \Rattle. This commit changed the script which generates the \texttt{osdef.h}, in such a way that it did not change the contents of \texttt{osdef.h}. \Make rebuilt all of the object files which depend on the header file, while \Rattle skipped them as the file contents had not changed.
\item For the 29th commit, \Rattle did much more work than \Make. This commit changed a prototype file which is indirectly included by all object files.  This prototype file is not listed as a dependency of the object files, so \Make did not rebuild them, but \Rattle noticed the changed dependency.  This missing dependency could have resulted in an incorrect build or a build error, but in this case (by luck) seems not to have been significant. A future prototype change may result in incorrect builds.
\end{enumerate}

% 21 is e258368; version.c and osdef.sh changed
% osdef.h is rebuilt; make says all object files depend on this so theyre all rebuilt, which rattle list says they do so; why did it not rebuild them? confusing debug code says osdef.h changed; but that doesn't seem to be the case since the object files that depend on it didnt say it had changed... i'll verify this

% 30 is a259d8d;
\subsection{tmux}
\label{sec:eval:tmux}

\begin{tikzpicture}
\begin{axis}[
  title=Compile time at each successive commit,
  width=\textwidth,
  height=5cm,
  ylabel={Seconds},
  ymax=50,
  ymin=0,
  xmin=0,
]

\addplot [color=cyan, mark=o] table [x expr=\coordindex, y=make1] {data/tmux.dat}; 	\addlegendentry{\Make -j1}
\addplot [color=cyan, mark=*] table [x expr=\coordindex, y=make4] {data/tmux.dat}; 	\addlegendentry{\Make -j4}
\addplot [color=purple, mark=triangle] table [x expr=\coordindex, y=rattle1_noshared] {data/tmux.dat}; 	\addlegendentry{\Rattle -j1}
\addplot [color=purple, mark=triangle*] table [x expr=\coordindex, y=rattle4_noshared] {data/tmux.dat}; 	\addlegendentry{\Rattle -j4}
\end{axis}
\end{tikzpicture}


tmux is a terminal multiplexer in development for the last 12 years.  It is built from source using a combination of \textsc{Autoconf}, \textsc{Automake}, \textsc{pkg-config} and \Make.  The build commands are:

\begin{verbatim}
sh autogen.sh
./configure
make
\end{verbatim}

To compare a \Rattle based tmux build to the \Make based one, we first ran \texttt{sh autogen.sh \&\& ./configure}, and then used the resulting \Make scripts to generate \Rattle build scripts. tmux was built over a series of 38 commits with both \Make and \Rattle. For the 11th, 22nd and 23rd commits \Rattle and \Make took noticeably different amounts of time. In all cases the cause was the build script changing, causing \Rattle to be unable to speculate, and thus taking approximately as long as a full single-threaded build.


% 47174f51
% 150 cmds in file - 10 for make
% 140 cmds run by rattle not run by make
% how many did rattle run period?
% there were 141 cmds in the script.
% so what command did make run too?

%4822130b was a new build script

% 685eb381 also a new build script



% For 1 thread

% commits where did same things: close enough; ignoring commands like rm tmux

% obf153da 0c6c8c4e 0eb7b547 19d5f4a0 22e9cf04 24cd726d 32be954b 37919a6b 3e701309
% 400750bb 43b36752 4694afb  470cba35 54553903 60ab7144 61b075a2 6c28d0dd 6f0241e6
% 74b42407 7cdf5ee9 7f3feb18 8457f54e 8b22da69 9900ccd0 a01c9ffc a4d8437b ba542e42
% bc36700d c391d50c c915cfc7 cdf13837 e9b12943 fdbc1116


% commits with differences
% 47174f51 (commit where make did automake stuff and rattle re-ran everything because a version number changed)
% 4822130b
% 685eb381

% 7eada28f: make did ./etc/ylwrap ....
% ed16f51e: make did ./etc/ylwrap ...
% ee3d3db3: make built tmux.o
% f3ea318a: make did ./etc/ylwrap ...


% commits
% ed16f51e 61b075a2 e9b12943 3e701309 8457f54e a01c9ffc cdf13837 74b42407 0eb7b547 f3ea318a 7cdf5ee9 ee3d3db3 685eb381 60ab7144 7eada28f 7f3feb18 8b22da69 bc36700d 32be954b 6f0241e6 19d5f4a0 43b36752 0bf153da 4822130b 47174f51 c915cfc7 54553903 400750bb 470cba35 a4d8437b 6c28d0dd 24cd726d 9900ccd0 c391d50c 0c6c8c4e fdbc1116 37919a6b 22e9cf04 ba542e42 4694afb

% \subsection{openssl}

\subsection{Node.js}
\label{sec:eval:node}

\begin{tikzpicture}
\begin{axis}[ymode=log,
  title=Compile time at each successive commit,
  width=\textwidth,
  height=5cm,
  ylabel={Seconds},
  xmin=0,
]

\addplot [color=cyan, mark=o] table [x expr=\coordindex, y=make1] {data/node.dat}; 	\addlegendentry{\Make -j1}
\addplot [color=cyan, mark=*] table [x expr=\coordindex, y=make4] {data/node.dat}; 	\addlegendentry{\Make -j4}
\addplot [color=purple, mark=triangle] table [x expr=\coordindex, y=rattle1_noshared] {data/node.dat}; 	\addlegendentry{\Rattle -j1}
\addplot [color=purple, mark=triangle*] table [x expr=\coordindex, y=rattle4_noshared] {data/node.dat}; 	\addlegendentry{\Rattle -j4}
\end{axis}
\end{tikzpicture}
% some data but not all; will put in tomorrow

% make4 2373.015 s
% rattle4 9491.027 s
% make1 9233.433 s
% rattle1 9189.8 s

% Brief into to the project
Node.js is a JavaScript runtime built on Chrome's V8 Javascript engine and can be found on \href{``https://github.com/nodejs/node''}{GitHub}.  The project is largely written in JavaScript, C++, Python, and C, and is built using Make and a meta-build tool called \emph{Generate Your Projects (GYP)}.  To build Node.js from source, a user first runs \emph{./configure} that configures the build and runs GYP.  GYP generates the majority of the Makefiles used to build the project.

GYP takes a series of \emph{.gyp} files and produces the project's build from those.  It puts all generated files in the project's \emph{out} directory and generates a separate \emph{*.mk} Makefile for each target, all of which are included by the top-level Makefile.  The generated top-level Makefile includes a \emph{do\_cmd} function which first checks if a command changed since the last time it was run, and if it has runs the command.  Additionally, it writes the command run to a generated dependency file, and if the command run was a \emph{g++} command that produced a file with dependencies, the dependencies in that file are cleaned up and written to the generated dependency file.

These dependency files are included in the top-level \emph{out/Makefile}, presumably so on future runs Make can consider the recorded dependencies as well as check whether the command has changed since it was previously run.  Most of these targets include the dependency \emph{FORCE\_DO\_CMD} which is an empty phony target and serves the purpose of forcing the build to check each time whether the command has changed since it was previously run. Each time a command is re-run, these dependency files are re-generated.

% Explain the makefile insanity and all of this dependency stuff they do
% Still trying to make coherent since of what this gyp is doing, so here is my current understanding

The build seems to both be using a form of tracing to keep track of accurate dependencies as well as tracking whether or not the build itself has changed.  \Rattle would make it unnecessary for the build to explicitly do this, since it internally traces and tracks the dependencies of all commands run, and doesn't need to worry about when a build changes.

% The generator/make.py

% what is this
% # Helper that is non-empty when a prerequisite changes.
% # Normally make does this implicitly, but we force rules to always run
% # so we can check their command lines.
% #   $? -- new prerequisites
% #   $| -- order-only dependencies
% prereq_changed = $(filter-out FORCE_DO_CMD,$(filter-out $|,$?))

% # do_cmd: run a command via the above cmd_foo names, if necessary.
% # Should always run for a given target to handle command-line changes.

% Explain the intermediate files and how on a rebuild with no changes stuff still happens

When building certain targets, \emph{.intermediate} files are created and at the end of the build deleted.  These \emph{.intermediate} commands appear to be for doing code generation.  So, the generated files depend on the \emph{.intermediate} target which runs the code generator.  As with the object file targets mentioned, a \emph{.intermediate.d} dependency file is generated for the \emph{.intermediate} target.  This dependency file contains the \emph{.intermediate} code generation command run, so the build can check whether the command changed since it was last run.

These \emph{.intermediate} files are specified as prerequisites of the special \Make target {.INTERMEDIATE}, which means that when one does not exist \Make won't bother updating it unless one of its prerequisites has changed.  Because these \emph{.intermediate} targets have \emph{DO\_FORCE\_CMD} as a prerequisite they always run.  This would enable the build to check if the command changed since it was last run, but this only works if the Makefile includes the dependency file storing that information.  The dependency files of these \emph{.intermediate} targets, which run every time, are not included in the Makefile.  Because of this the Makefile thinks the command has never run before, causing them to run every time.  Even if the Makefile is modified to include the dependency files of these \emph{.intermediate} targets, the targets still execute the recipe which does the file generation even though the command has not changed. % SS comment on this further when I fully understand why the commands are registering as changed...

  % Other things like parallelism and rattle rebuilding less

It appears the authors of this build are attempting to keep track of accurate dependencies by recording them every time a command runs, they are also tracking the build by recording every command run and checking whether a command or its prerequisites have changed when deciding whether to run it.  The scheme to accomplish this is in my opinion quite complicated and confusing and relies on various hacks to work around \Make and the fact that this isn't a full-fledged programming language.  \Rattle accomplishes both of these things the build authors are trying to accomplish with \Make without forcing them to work so hard.

To compare the \Make based build to a \Rattle version, we built Node.js over a series of commits checked-out from GitHub with both the original build system and the adapted \Rattle build.  The \Rattle build was created by recording every command executed by the original build, except for the commands creating dependency files.  A new such \Rattle build script was generated for any commit where the original build system changed or files were added to or removed from the project.

Node was built over a series of 38 commits with both the original \Make build and the adapted \Rattle build, with 1 thread and 4 threads.  \Make and \Rattle took approximately the same amount of time when building Node from scratch with a single thread.  Since \Rattle always does the first build sequentially, with 4 threads \Make obviously built Node more quickly.  For a substantial number of the commits \Rattle was faster than the \Make based build.  As we can see from the chart, there is a baseline of time that the \Make build never dips below.  This is because as mentioned above, there are a series of targets in the \Make build that are set up in such a way that they always run.  \Rattle avoids rebuilding these targets on every build.  This is seems to be an error on the part of Node's build authors and just goes to show how difficult it is to get build systems and dependencies correct.

% SS Note to Neil: I need to add note about hazards; can see in chart where 4core is slower than 1 core that hazard probably ahppened.

% SS add the differences in commits
% 1. make list of commits run

% in order from latest to oldest.
% d80c40047b 0fe810168b 22724894c9 ab9e89439e cb210110e9 d10927b687
% 023ecbccc8 be6596352b [54c1a09202] 470511ae78 25c3f7c61a 13fe56bbbb
% [43fb6ffef7] a171314003 dd4c62eabe abe6a2e3d1 9225939528 d4c81be4a0
% 38aa31554c 1d9511127c d227d22bd9 5cf789e554 d65e6a5017 24e81d7c5a
% 2cd9892425 [64161f2a86] 0f8941962d 2170259940 32f63fcf0e 2462a2c5d7
% b851d7b986 70c32a6d19 3d456b1868 f2ec64fbcf 59a1981a22 [7b7e7bd185]
% 78743f8e39 a5d4a397d6
% [] indicate a commit where build appears to have changed. so need new rattle build too


% Comment from the Makefile
%# .buildstamp needs $(NODE_EXE) but cannot depend on it
%# directly because it calls make recursively.  The parent make cannot know
%# if the subprocess touched anything so it pessimistically assumes that
%# .buildstamp is out of date and need a rebuild.
%# Just goes to show that recursive make really is harmful...
%# TODO(bnoordhuis) Force rebuild after gyp update.
